Nonlinear Tidal Dissipation in Binary White Dwarfs

### 2019-01-27 proposal/FINESST/Su_FINESST_2019.tex
update plots, draft v2 of FINESST proposal

\section{White Dwarf Binaries}

Compact white dwarf (WD) binary systems, with orbital periods in the range of
minutes to hours, are important for several topics in astrophysics. They are
expected to be important sources of gravitational wave (GW) radiation for the
Laser Interferometric Space Antenna (LISA)\cite{lisa}. They are also thought to
be likely progenitors of type Ia supernovae (e.g.\ \cite{Ia0,webbink} or more
recently\cite{Ia1,Ia2}). While presently only a few tens of compact WD binaries are
known, \emph{Gaia} (currently gathering data) is expected to expand the catalog
to a few hundred, and the Large Synoptic Survey Telescope (LSST, first light
scheduled for 2020) is expected to detect a few thousand more\cite{lsst_wd}.
These observations are expected to significantly advance understanding of WD
binaries: results based on \emph{Gaia}'s second data release have already begun
to appear\cite{gaiaDD,gaiaDD2}. The proposed research is well-timed to take
advantage of these new resources.

In spite of the broad importance of WD binaries, the evolution of these systems
pre-merger is sparsely studied. In particular, compact WD binaries are expected
to exhibit major tidal interactions. In these interactions, tidal forces from
the companion WD excite internal gravity waves (IGW), displacements in the WD
fluid experiencing a buoyant restoring force due to density stratification.
As these waves propagate outwards from where they are excited, they grow in
amplitude until they break, as do ocean waves on a shore, and deposit both
energy and angular momentum from the binary orbit in the outer envelope of the
WD\@.

Previous works have shown that this tidal dissipation mechanism can generate
significantly more energy than thermal radiation from the WD surface alone and
is thus a major contributor to the WD energy budget\cite{fullerII,fullerIV}.
However, these works treated the nonlinear wave breaking in a parameterized
fashion. Related works in other fields find that properly including nonlinear
phenomena in hydrodynamical wave breaking produces drastically different IGW
dissipation behavior\cite{winters1994,barker_ogilvie}. Such fully nonlinear
studies have not been performed for WD binaries but are important for the
thermal and orbital evolution of WDs undergoing tidal heating. Characterizing
this nonlinear tidal dissipation will require numerical simulation to capture
the turbulent cascade to small scales that drives IGW dissipation.

\textbf{I propose to study the dynamical effects of tidal dissipation via
nonlinear IGW breaking in binary WDs.} To accomplish this, I aim to address the
following three aims:
\begin{itemize}
    \item Characterize the location, spatial extent, and other properties of
        wave breaking in realistic WD models via direct numerical simulation.
        The location and spatial extent will furnish a simple yet effective
        parameterization of tidal dissipation that can substitute full
        hydrodynamic simulation.

    \item Predict signatures of tidal dissipation over a wide range of possible
        WD systems. In particular, I will describe the impact of tidal heating
        on the apparent temperature of a binary WD and maybe even the production
        of observable flares.

    \item Compute modified GW templates for LISA that account for changes in the
        phase evolution of the orbit due to tidal dissipation.
\end{itemize}

\section{Nonlinear Tidal Dissipation}

@@

The current understanding of tidal synchronization in WD binaries is laid out
in~\cite{fullerII}: a tidally-excited train of IGW undergoes wave breaking
in the outer envelope of the WD, locally depositing angular momentum and
synchronizing the WD spin to the orbit. Such a process is broadly consistent
with canonical models of tidal synchronization in stars\cite{zahn75,gn89}, the
only major difference being in the specifics of wave excitation\footnote{While
IGWs in stars are excited at their radiative-convective boundaries, IGWs in WDs
are excited at sharp changes in composition.}. While this tidal synchronization
mechanism is generally accepted, direct numerical simulation of the wave
breaking process is sparse. Since wave breaking is a strongly nonlinear
phenomenon, where a larger wave breaks down into many smaller-scale waves,
numerical simulation is paramount to an accurate understanding of IGW breaking.

IGW breaking is well studied in atmospheric sciences. The wave breaking process
proceeds as follows: initially, the universal instability of an IGW to the
parametric subharmonic instability is responsible for transfering energy and
angular momentum from the unstable wave to the mean flow of the
fluid\cite{drazin}. Subsequently, after the mean flow velocity reaches the
horizontal phase velocity of the IGW, a well-known calculation shows that the
IGW is nearly completely absorbed by the mean flow in the linear
approximation and endowing the atmosphere with a mean horizontal
flow\cite{booker_bretherton,hazel}. However, when this mean flow absorption was
numerically studied including full nonlinear interactions, \emph{new phenomena
not described by the linear theory were observed}\cite{jones_num,winters1994}.
This highlights the importance of numerical simulation in capturing the wave
breaking process.

To this end, I have adapted the spectral hydrodynamic code Dedalus\cite{dedalus}
to simulate a plane-parallel IGW wavetrain excited at the bottom of a 2D
incompressible fluid. A spectral code like Dedalus is ideal for simulating
complex hydrodynamical phenomena as spectral methods have no inherent numerical
viscosity and so are more suitable for resolving the nonlinear cascade to small
length scales in wave breaking.

In my simulations, I observe IGW propagating upwards from where they are
excited, then breaking and depositing horizontal momentum in the fluid, causing
the fluid to acquire average horizontal momentum, consistent with
literature\cite{fullerII}. I derive simple formulae for the location and spatial
extent of the dissipation zone where the IGW is absorbed by the fluid. More
interestingly, I observe partial reflection of the IGW at the synchronization
layer\cite{me}, a phenomenon not considered in the current astrophysical
literature but consistent with atmospheric science results\cite{winters1994}. A
sample simulation is presented in \autoref{fig:nl_fluxes}. I am preparing these
results for publication\cite{me}.
\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth]{nl_fluxes.png}
    \caption{The evolution of the horizontal momentum flux $S_{px}(z)$ over time
    in a representative simulation. Note that the location where horizontal
    momentum is deposited by wave breaking, the sharp decrease in $S_{px}$,
    moves to lower $z$ at later times; this reflects spinup of the fluid from
    the top down. The incident flux appears to decrease over time as well, but
    this is attributed to partial reflection of the IGW off the critical
    layer. The vertical axis is in units of the total excited horizontal momentum
    flux, while distances and time are in units of the stratification scale
    height and the buoyancy period.}\label{fig:nl_fluxes}
\end{figure}

@@

    \item I will extend my simulations to realistic WD models and equations of
        state such as those in~\cite{brassard1992}, continuing to track the
        location and spatial extent of the dissipation layer as well as any new
        phenomena. As WDs vary widely in composition and effective temperature,
        studying representative WD models is vital to obtaining a robust
        characterization of tidal dissipation. My selected WD models are those
        used in the current literature\cite{fullerII, fullerIV}.

    \item With these simulations, I will build a model of tidal dissipation
        valid for a broad range of WDs. I will package my model as a module for
        the 1D stellar evolution code Modules for Experiments in Stellar
        Astrophysics (MESA)\cite{MESA}. MESA is a popular choice for stellar
        evolution simulation due to its computational speed and extensibility.
        It has been used in previous studies to study the evolution of WDs under
        tidal heating\cite{fullerIV,tidal_novae}. My module will provide a
        parameterized and numerically-validated model of tidal dissipation that
        can be used to quickly and easily model tidal interactions in binary
        WDs.
\end{itemize}

\section{Tidal Heating Thermodynamics}

@@

With the tidal dissipation models I will develop, I will be able to perform a
simulation similar to that in FL but with a more realistic tidal interaction.
The second aim of my proposal is accordingly: \textbf{I will use my tidal
dissipation profiles to simulate a binary WD undergoing tidal heating and make
comparisons to observational data.} The location of the tidal heating is a key
ingredient in determining whether deposited energy can be efficiently transfered
away or whether the WD experiences sudden detonations. As such, an improved
understanding of the properties of tidal dissipation is important to proper
forecasting of binary WDs' thermodynamic evolution. Moreover, while only a few
sufficiently compact WD binaries were available at the time of writing of FL,
\emph{Gaia} data releases 2 and 3 will provide many more compact WD binaries to
examine for evidence of tidal heating.

@@

\section{Tidal Heating and LISA}

@@

During the first year of work and first half of the second year, I will complete
my tidal dissipation model. I anticipate the extension of my 2D plane-parallel
work to more complex geometries will be complete within the first half year,
while the extension to realistic WD models will take up to a year. I expect that
these two results together will produce one peer-reviewed publication and one
MESA module.

>>>>

\section{Background and Proposed Research}

\subsection{White Dwarf Binaries}

Compact white dwarf (WD) binary systems, with orbital periods in the range of
minutes to hours, are important for a range of astrophysical problems. They are
the most important sources of gravitational waves (GWs) for the Laser
Interferometric Space Antenna (LISA)\cite{lisa}. They are also thought to
produce interesting optical transients such as tidal novae\cite{tidal_novae},
underluminous supernovae\cite{underlum}, and Ca-rich fast
transients\cite{carich}. Most importantly, they have been proposed as the likely
progenitors of type Ia supernovae (e.g.~\cite{Ia0,webbink} or more
recently\cite{Ia1,Ia2}). While presently only a few tens of compact WD binaries
are known\cite{lsst_wd}, \emph{Gaia} (currently gathering data) is expected to
expand the catalog to a few hundred\cite{lsst_wd} (results based on
\emph{Gaia}'s second data release have already begun to
appear\cite{gaiaDD,gaiaDD2}), and the Large Synoptic Survey Telescope (LSST,
first light scheduled for 2020) will likely detect a few thousand
more\cite{lsst_wd}. These observations will significantly advance the
understanding of WD binaries. My proposed theoretical and computational research
is well-timed to take advantage of these new advances.

In spite of the broad importance of WD binaries, the evolution of these systems
prior to their final merger is not well understood. Much of this uncertainty
comes from our imprecise understanding of tidal interactions, which play an
important role during a compact WD binary's inspiral\cite{fullerII}. Studies
show that these interactions manifest as tidal excitation of internal gravity
waves (IGW), waves in the WD fluid restored by the buoyancy force due to density
stratification\cite{fullerI}. As these waves propagate outwards towards the
surface, they grow in amplitude until they break, as do ocean waves on a shore,
and transfer both energy and angular momentum from the binary orbit to the outer
envelope of the WD\cite{fullerI,fullerII}.

Previous works have found that this tidal dissipation mechanism can generate
significantly more energy than thermal radiation from the WD surface alone and
is thus a major contributor to the WD energy budget\cite{fullerII,fullerIV}.
However, these works parameterized the wave breaking process without
hydrodynamical simulations. The details of tidal dissipation, namely the
location and spatial extent of the wave breaking, affect the observable outcome:
dissipation near the surface of the WD can be efficiently radiated away and
simply brightens the WD, while dissipation deep in the WD causes an energy
buildup that results in energetic flares\cite{tidal_novae}. Similar works in
other fields find that numerical simulations capturing the strongly nonlinear
wave breaking process find new behavior that cannot be described by linear and
weakly nonlinear theory\cite{winters1994,barker_ogilvie}. Such fully nonlinear
numerical simulations have not been performed for WDs.

\subsection{Proposed Research}

Characterizing the location and spatial extent of tidal dissipation in WD
binaries will require numerical simulation to capture the turbulent cascade to
small scales causing wave breaking. \textbf{I propose to study the dynamical
effects of tidal dissipation via nonlinear IGW breaking in binary WDs.} To
accomplish this, I aim to address the following three aims:
\begin{itemize}
    \item Characterize the location, spatial extent, and other properties of
        wave breaking in realistic WD models via direct numerical simulation.
        The location and spatial extent will furnish a simple yet effective
        parameterization of tidal dissipation. \autoref{s:2} describes the
        steps required for this aim.

    \item Predict signatures of tidal dissipation over a wide range of possible
        WD systems. In particular, I will describe the impact of tidal heating
        on the apparent temperature of a binary WD and maybe even the production
        of observable flares. \autoref{s:3} details my plan to do so.

    \item Compute modified GW templates for LISA that account for changes in the
        phase evolution of the orbit due to tidal dissipation. \autoref{s:4}
        elaborates on how I will perform this computation.
\end{itemize}

\section{Nonlinear Tidal Dissipation}\label{s:2}

@@

The current understanding of tidal synchronization in WD binaries is laid out
in~\cite{fullerII}: a tidally-excited train of IGW undergoes wave breaking in
the outer envelope of the WD, locally depositing angular momentum and
synchronizing the WD spin to the binary orbit. A similar process may also
operate in stellar binaries consisting of early type stars\cite{zahn75,gn89},
the only major difference being in the specifics of wave
excitation\footnote{While IGWs in stars are excited at their
radiative-convective boundaries, IGWs in WDs are excited at sharp changes in
composition.}. Nevertheless, direct numerical simulation of the wave breaking
process has not been performed in either of these systems. Since wave breaking
is a strongly nonlinear phenomenon, where a larger wave breaks down into many
smaller-scale waves, numerical simulation is paramount to an accurate
understanding of IGW breaking.

IGW breaking has been studied in detail in atmospheric sciences. The wave
breaking process proceeds as follows: initially, as the IGW reaches nonlinear
amplitudes, it breaks down via the parametric subharmonic instability and
transfers energy and angular momentum from the wave to the mean flow of the
fluid\cite{drazin}. Subsequently, after the mean flow velocity reaches the
horizontal phase velocity of the IGW, a critical layer through which the IGW
cannot propagate forms. A well-known calculation shows that the IGW is nearly
completely absorbed at this critical layer in the linear approximation and
endows the atmosphere with a mean horizontal flow\cite{booker_bretherton,hazel}.
However, when this mean flow absorption was numerically studied including full
nonlinear interactions, new phenomena not described by the linear theory
(reflection off the critical layer and sharpening of the mean flow) were
observed\cite{jones_num,winters1994} that affected the evolution of the
atmosphere over time. \emph{This highlights the importance of numerical
simulation in capturing the wave breaking process.}

To gain insight into the tidal dissipation process, I began by using the
spectral hydrodynamics code Dedalus\cite{dedalus} to study IGW breaking in a 2D
isothermal, stratified atmosphere. A spectral code like Dedalus is ideal for
simulating complex hydrodynamical phenomena, as spectral methods have no
inherent numerical viscosity and so better resolve the nonlinear cascade to
small length scales in wave breaking.

Working with Dr.\ Daniel Lecoanet (Princeton; one of the authors of the Dedalus
code) and my advisor, Prof.\ Dong Lai, I simulated an upward-propagating IGW
wavetrain excited at the bottom of the atmosphere. I observe the excited wave
breaking and depositing horizontal momentum in the fluid, causing the fluid to
acquire an average horizontal flow, consistent with previous
studies\cite{fullerII}. I derive simple formulae for the location and spatial
extent of the dissipation zone where the IGW is absorbed by the fluid. More
interestingly, I observe partial reflection of the IGW at the synchronization
layer\cite{me}, a phenomenon not considered in the current astrophysical
literature but consistent with the aforementioned results\cite{winters1994}. A
sample simulation is presented in \autoref{fig:nl_fluxes}. I am preparing these
results for publication\cite{me}.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth]{nl_fluxes.png}
    \caption{The evolution of the average horizontal flow velocity of the fluid
    $U_0$ (in units of $c_{ph,x}$ the horizontal phase velocity of the IGW) and
    horizontal momentum flux contained in the IGW $S_{px}(z)$ (in units of the
    total excited flux $S_0$) for one of my simulations. Different colored lines
    correspond to different times in the simulation. Times are measured in units
    of $N^{-1}$, the inverse of the Brunt-V\"ais\"al\"a frequency (or the
    buoyancy period), and heights in units of $H$ the scale height of the
    density stratification. IGW are excited at $z = 2H$ and propagate to higher
    $z$ before undergoing wave breaking. The sharp decrease in $S_{px}$ is
    the location of the critical layer where the IGW breaks and drives $U_0$
    towards $c_{ph, x}$; it moves to lower $z$ over time. The apparent decrease
    in $S_{px}$ at later times indicates reflection off the critical
    layer.}\label{fig:nl_fluxes}
\end{figure}

@@

    \item I will extend my simulations to realistic WD models and equations of
        state such as those in~\cite{brassard1992}, continuing to track the
        location and spatial extent of the dissipation layer as well as any new
        phenomena. As WDs vary widely in composition and effective temperature,
        studying representative WD models is vital to obtaining a robust
        characterization of tidal dissipation. My selected WD models are those
        used in the current literature\cite{fullerII, fullerIV}.
\end{itemize}

\section{Tidal Heating Thermodynamics}\label{s:3}

@@

With the tidal dissipation models I will develop, I will be able to perform a
simulation similar to that in FL but with a more realistic tidal interaction.
The second aim of my proposal is accordingly: \textbf{I will use my tidal
% "obtained in Section 2"
dissipation profiles to simulate a binary WD undergoing tidal heating and make
comparisons to observational data.} The location of the tidal heating is a key
ingredient in determining whether deposited energy can be efficiently transfered
away or whether the WD experiences sudden detonations. As such, an improved
understanding of the properties of tidal dissipation is important to proper
forecasting of binary WDs' thermodynamic evolution. Moreover, while only a few
sufficiently compact WD binaries were available at the time of writing of FL,
\emph{Gaia} data releases 2 and 3 will provide many more compact WD binaries to
examine for evidence of tidal heating.

@@

\section{Tidal Heating and LISA}\label{s:4}

@@

During the first year of work and first half of the second year, I will complete
my tidal dissipation model. I anticipate the extension of my 2D plane-parallel
work to more complex geometries will be complete within the first half year,
while the extension to realistic WD models will take up to a year. I expect that
these two results together will produce one peer-reviewed publication in
addition to the one currently in preparation.
